---
title: "`r yaml::read_yaml('config.yaml')$project$name`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    self_contained: true
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
# Function import
function_files <- list.files("Functions", pattern = "\\.R$", full.names = TRUE)
invisible(lapply(sort(function_files), source))

config <- read_yaml("config.yaml")
create_project_dirs(overwrite = config$run$overwrite)

# Sets status for pipeline modules
run_env <- isTRUE(config$analysis$run_env)
run_qc <- isTRUE(config$analysis$run_qc)
run_ld_pruning <- isTRUE(config$analysis$run_ld_pruning)
run_pca <- isTRUE(config$analysis$run_pca)
run_kinship <- isTRUE(config$analysis$run_kinship)
run_lmm <- isTRUE(config$analysis$run_lmm)
run_lfmm <- isTRUE(config$analysis$run_lfmm)
run_rda <- isTRUE(config$analysis$run_rda)
run_pcadapt <- isTRUE(config$analysis$run_pcadapt)
run_consensus <- isTRUE(config$analysis$run_consensus)
run_consensus_ld <- isTRUE(config$analysis$run_consensus_ld)
run_consensus_r <- isTRUE(config$analysis$run_consensus_robustness)
run_adaptive_scoring <- isTRUE(config$analysis$run_adaptive_scoring)
run_bio_interp <- isTRUE(config$analysis$run_bio_interp)
```

# Environmental Data Processing
## Extracting climatic variables
```{r, eval = run_env, include = run_env}
# Extracts climate variables

# if filtering by samples
if (config$subsetting$use) {
  # Reads metadata for subset
  metadata <- read.csv(config$subsetting$sample_metadata, stringsAsFactors = FALSE)
  # Extracts climatic variables for subset
  climate_data <- extract_climate_variables(
  metadata = metadata,
  lon_col = config$env$lon_col,
  lat_col = config$env$lat_col,
  sample_col = config$metadata$sample_col,
  var = config$env$var,
  res = config$env$res,
  climate_dir = config$env$path,
  output_file = config$env$output_file,
  overwrite = config$run$overwrite
  )
} else {
  metadata <- read.csv(config$metadata$file, stringsAsFactors = FALSE)
  climate_data <- extract_climate_variables(
  metadata = metadata,
  lon_col = config$env$lon_col,
  lat_col = config$env$lat_col,
  sample_col = config$metadata$sample_col,
  var = config$env$var,
  res = config$env$res,
  climate_dir = config$env$path,
  output_file = config$env$output_file,
  overwrite = config$run$overwrite
  )
}
```

## Climatic Data Distribution
### Geographic distribution
```{r, eval = run_env, include = run_env}
# Plots geographic distribution of selected climate variables
for (var in config$env$vars) {
  print(
    plot_climate_map(
      climate_data = climate_data,
      metadata = metadata,
      sample_col = config$metadata$sample_col,
      lon_col = config$env$lon_col,
      lat_col = config$env$lat_col,
      color_col = var
    )
  )
}
```

### Correlation
```{r, eval = run_env, include = run_env}
# Plots a correlation heatmap of climate variables
plot_climate_cor_heatmap(
  climate_data = climate_data
)
```

### Climate PCA
```{r, eval = run_env, include = run_env}
# Runs PCA of climate variables
climate_pca <- run_climate_pca(
  climate_data = climate_data,
  output_file = "Output/Structure/climate_pca.rds"
)

# Plots PCA
plot_climate_pca(climate_pca)
```

# Genotype Quality Control
```{r, eval = run_qc, include = run_qc}
qc_files <- paste0(config$qc_outputs$qc_prefix, c(".bed", ".bim", ".fam"))

# Filters genotype data to match samples with climatic variable data
keep_file <- create_keep_file(
  metadata = climate_data,
  fam_file = paste0(config$genotype$input_prefix, ".fam"),
  output_file = config$subsetting$keep_file,
  sample_col = config$metadata$sample_col
)

# if overwrite = FALSE and filtered files exist
if (!config$run$overwrite && all(file.exists(qc_files))) {
  message("Reusing filtered PLINK files: ", config$qc_outputs$qc_prefix)
  qc_prefix <- config$qc_outputs$qc_prefix
} else {
  # QC filters using PLINK
  qc_prefix <- plink_filter(
    input = config$genotype$input_prefix,
    output = config$qc_outputs$qc_prefix,
    plink = config$plink$path,
    keep = keep_file,
    geno_na = config$qc$geno_na,
    maf = config$qc$maf,
    ind_na = config$qc$ind_na
  )
}

# Extracts markers use_extract = TRUE
if (isTRUE(config$marker_subsetting$use_extract)) {
  qc_prefix <- plink_extract_markers(
    input = qc_prefix,
    output = config$marker_subsetting$output_prefix,
    plink = config$plink$path,
    extract_file = config$marker_subsetting$extract_file,
    overwrite = config$run$overwrite
  )
}

# Refilters metadata to match genotype data
metadata <- filter_metadata(
  metadata_file = config$metadata$file,
  fam_file = paste0(qc_prefix, ".fam"),
  sample_col = config$metadata$sample_col,
  output_file = config$metadata$filtered_file
)
```

```{r}
# In case QC module is not ran
if (!exists("qc_prefix")) {
  qc_prefix <- config$qc_outputs$qc_prefix
  qc_files <- paste0(qc_prefix, c(".bed", ".bim", ".fam"))
  # if files do not exist
  if (!all(file.exists(qc_files))) {
    message("QC was not run and QC files were not found. Using raw genotype prefix.")
    qc_prefix <- config$genotype$input_prefix
  } else {
    message("QC was not run. Reusing existing QC files: ", qc_prefix)
  }
}
```

```{r}
qc_plots <- plot_genotype_qc_overview(
  prefix = qc_prefix,
  output_dir = "Output/QC",
  dataset_name = "QC_genotype"
)

qc_plots$marker_summary %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)

qc_plots$marker_count_plot

qc_plots$marker_position_plot
```


# Population Structure & Kinship
## LD-pruning
```{r, eval = run_ld_pruning, include = run_ld_pruning}
ld_files <- paste0(config$qc_outputs$ld_prefix, c(".bed", ".bim", ".fam"))

# if overwrite = FALSE and ld_files exist
if (!config$run$overwrite && all(file.exists(ld_files))) {
  message("Reusing existing LD-pruned PLINK files: ", config$qc_outputs$ld_prefix)
  ld_prefix <- config$qc_outputs$ld_prefix
} else {
  # Filters genotype data with LD
  ld_prefix <- plink_ld_prune(
    input = qc_prefix,
    output = config$qc_outputs$ld_prefix,
    plink = config$plink$path,
    window = config$ld_pruning$window,
    step = config$ld_pruning$step,
    r2 = config$ld_pruning$r2
  )
}
```

## PCA
```{r, eval = run_pca, include = run_pca}
# Loading LD-pruned PLINK data as bigSNP object
ld_obj <- plink_to_bigSNP(
  bed_file = paste0(ld_prefix, ".bed"),
  overwrite = config$run$overwrite
)

# Runs PCA on genotype data
pca_result <- run_pca_bigsnpr(
  obj = ld_obj,
  n_pcs = config$pca$n_pcs,
  output_file = config$pca$output_file,
  covariates_file = config$pca$covariates_file,
  ncores = config$pca$ncores
)

# Plots PCA
pca_plot <- plot_pca(
  pca_result = pca_result,
  metadata = metadata,
  sample_col = config$metadata$sample_col,
  color_col = NULL,
  pc_x = 1,
  pc_y = 2
)

pca_plot
```

## Kinship

```{r, eval = run_kinship, include = run_kinship}
# Creates GEMMA kinship matrix
gemma_kinship_file <- run_gemma_kinship(
  gemma = config$gemma$path,
  bfile = qc_prefix,
  output_prefix = config$gemma$kinship_prefix,
  output_dir = config$gemma$output_dir,
  overwrite = config$run$overwrite
)

# Calculates kinship statistics
kinship_stats <- summarize_gemma_kinship(gemma_kinship_file)

kinship_stats %>%
  mutate(value = round(value, 4)) %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

# GEA
## LMM (GEMMA)
```{r, eval = run_lmm, include = run_lmm}
# Defines result files
gemma_files <- file.path(
  config$gemma$output_dir,
  c("gemma_all_results.csv",
    "gemma_all_evaluation.csv",
    "gemma_best_strategy_by_var.csv")
)

if (!config$run$overwrite && all(file.exists(gemma_files))) {
  message("Reusing existing GEMMA results.")
  gemma_results <- load_gemma_results(config)
} else {
  # Runs GEMMA
  gemma_results <- run_gemma_lmm_all_variables(
    config = config,
    qc_prefix = qc_prefix,
    climate_data = climate_data,
    kinship_file = gemma_kinship_file,
    phenotypes = config$env$vars
  )
}
```

### Results
#### Evaluation
```{r, eval = run_lmm, include = run_lmm}
gemma_results$evaluation %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

#### Best by variable
```{r, eval = run_lmm, include = run_lmm}
gemma_results$best_by_variable %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

### QQ Plots
```{r, eval = run_lmm, include = run_lmm}
for (pheno in gemma_results$best_by_variable$phenotype){
  print(
    plot_gemma_qq(
      results = gemma_results$results,
      phenotype = pheno
      )
  )
}
```

## LFMM
```{r, eval = run_lfmm, include = run_lfmm}
# Defines result files
lfmm_files <- file.path(
  config$lfmm$output_dir,
  c("lfmm_all_results.csv",
    "lfmm_all_evaluation.csv",
    "lfmm_best_strategy_by_var.csv")
)

if (!config$run$overwrite && all(file.exists(lfmm_files))) {
  message("Reusing existing LFMM results.")
  lfmm_results <- load_lfmm_results(config)
} else {
  # Creates genotype matrix for LFMM
  qc_obj <- plink_to_bigSNP(
    bed_file = paste0(qc_prefix, ".bed"),
    overwrite = config$run$overwrite
  )
  
  G_qc <- qc_obj$genotypes
  map_qc <- qc_obj$map
  fam_qc <- qc_obj$fam
  
  # Runs LFMM
  lfmm_results <- run_lfmm_all_variables(
    config = config,
    geno = G_qc,
    map = map_qc,
    fam = fam_qc,
    climate_data = climate_data,
    phenotypes = config$env$vars
  )
  
  rm(qc_obj, G_qc, map_qc, fam_qc)
  gc()
}
```

### Results
#### Evaluation
```{r, eval = run_lfmm, include = run_lfmm}
lfmm_results$evaluation %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

#### Best by variable
```{r, eval = run_lfmm, include = run_lfmm}
lfmm_results$best_by_variable %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

### QQ Plots
```{r, eval = run_lfmm, include = run_lfmm}
for (pheno in lfmm_results$best_by_variable$phenotype) {
  print(
    plot_lfmm_qq(
      results = lfmm_results$results,
      phenotype = pheno
    )
  )
}
```

## RDA
```{r, eval = run_rda, include = run_rda}
# Defines result files
rda_files <- file.path(
  config$rda$output_dir,
  c("rda_all_results.csv",
    "rda_all_evaluation.csv",
    "rda_best_strategy_by_var.csv")
)

# Use LD-pruned genotype data for RDA
if (!exists("ld_prefix")) {
  ld_prefix <- config$qc_outputs$ld_prefix
}

if (!config$run$overwrite && all(file.exists(rda_files))) {
  message("Reusing existing RDA results.")
  rda_results <- load_rda_results(config)
} else {
  # Creates genotype matrix for RDA
  qc_obj <- plink_to_bigSNP(
    bed_file = paste0(ld_prefix, ".bed"),
    overwrite = config$run$overwrite
  )
  
  G_qc <- qc_obj$genotypes
  map_qc <- qc_obj$map
  fam_qc <- qc_obj$fam
  
  # Runs RDA
  rda_results <- run_rda_all_variables(
    config = config,
    geno = G_qc,
    map = map_qc,
    fam = fam_qc,
    climate_data = climate_data,
    phenotypes = config$env$vars
)
  rm(qc_obj, G_qc, map_qc, fam_qc)
  gc()
}
```

### Results
#### Evaluation
```{r, eval = run_rda, include = run_rda}
rda_results$evaluation %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

#### Best by variable
```{r, eval = run_rda, include = run_rda}
rda_results$best_by_variable %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

### QQ Plots
```{r}
for (pheno in rda_results$best_by_variable$phenotype) {
  print(
    plot_rda_qq(
      results = rda_results$results,
      phenotype = pheno
    )
  )
}
```

## pcadapt
```{r, eval = run_pcadapt, include = run_pcadapt}
# Defines result files
pcadapt_files <- file.path(
  config$pcadapt$output_dir,
  c("pcadapt_all_results.csv",
    "pcadapt_all_evaluation.csv",
    "pcadapt_best_strategy.csv")
)

if (!config$run$overwrite &&
    all(file.exists(pcadapt_files))) {
  message("Reusing existing pcadapt results.")
  pcadapt_results <- load_pcadapt_results(config)
} else {
  # Run pcadapt
  pcadapt_results <- run_pcadapt_strategies(
    config = config,
    qc_prefix = qc_prefix
  )
}
```

### Results
```{r, eval = run_pcadapt, include = run_pcadapt}
pcadapt_results$evaluation %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)

pcadapt_results$best_evaluation %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

### QQ Plot
```{r}
plot_pcadapt_qq(
  results = pcadapt_results$results
)
```

# Consensus SNP sets
## Construction
```{r, eval = run_consensus, include = run_consensus}
# if overwrite = FALSE and consensus set files exist for all phenotypes
if (!config$run$overwrite && consensus_files_exist(config, phenotypes = config$env$vars)) {
  message("Reusing existing consensus SNP results.")
  consensus_results <- load_consensus_results(
    config = config,
    phenotypes = config$env$vars
  )
} else {
  # Create consensus SNP sets
  consensus_results <- run_consensus_all_variables(
    config = config,
    gemma_results = gemma_results,
    lfmm_results = lfmm_results,
    rda_results = rda_results,
    pcadapt_results = pcadapt_results,
    phenotypes = config$env$vars
  )
}
```

## Evaluation
```{r, eval = run_consensus, include = run_consensus}
# Consensus set evaluation
consensus_results$evaluation %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)

# Consensus sets for each variable
consensus_sets <- tagList(
  lapply(config$env$vars, function(phenotype) {
    res <- consensus_results$by_variable[[phenotype]]
    
    tagList(
      tags$h2(phenotype),
      
      tags$h3("Broad consensus"),
      make_reactable(res$broad_2methods),
      
      tags$h3("Environmental consensus"),
      make_reactable(res$env_2methods),
      
      tags$h3("High-confidence consensus"),
      make_reactable(res$high_confidence),
      
      tags$hr()
    )
  })
)

consensus_sets
```

# Post-Hoc LD Block Pruning
```{r, eval = run_consensus_ld, include = run_consensus_ld}
# if overwrite = FALSE and LD block files exist
if (!config$run$overwrite && ld_block_files_exist(config, phenotypes = config$env$vars)) {
  message("Reusing existing post hoc LD block pruning results.")
  # Read results from files
  ld_results <- load_ld_block_results(
    config = config,
    phenotypes = config$env$vars
  )
} else {
  # Run LD pruning and blocking on all variables
  ld_results <- run_ld_blocks_all_variables(
    config = config,
    qc_prefix = config$qc_outputs$qc_prefix,
    consensus_results = consensus_results,
    phenotypes = config$env$vars
  )
}
```

## Results
```{r, eval = run_consensus_ld, include = run_consensus_ld}
ld_results$summary %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

# Robustness Evaluation
```{r, eval = run_consensus_r, include = run_consensus_r}
# Runs robustness evaluation
robustness_results <- run_consensus_robustness(
  ld_results = ld_results,
  output_dir = config$consensus_robustness$output_dir
)
```

## Evaluation Results
```{r, eval = run_consensus_r, include = run_consensus_r}
robustness_results$evaluation %>%
  dplyr::select(-lead_snp_signature) %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

## Selected Primary Sets Evaluation Summary
```{r, eval = run_consensus_r, include = run_consensus_r}
robustness_results$primary_summary %>%
  dplyr::select(-lead_snp_signature) %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```

## Primary Lead SNPs (All Variables)
```{r, eval = run_consensus_r, include = run_consensus_r}
robustness_results$primary_lead_snps %>%
  reactable(sortable = TRUE, filterable = TRUE, searchable = TRUE)
```


# Adaptive Scoring
```{r, eval = run_adaptive_scoring, include = run_adaptive_scoring}
# Calculate adaptive score (for each variable) for all accessions
adaptive_results <- run_adaptive_germplasm_scoring(
  robustness_results = robustness_results,
  gemma_results = gemma_results,
  rda_results = rda_results,
  lfmm_results = lfmm_results,
  qc_prefix = qc_prefix,
  metadata = metadata,
  output_dir = config$adaptive_scoring$output_dir,
  sample_col = config$metadata$sample_col,
  overwrite = FALSE
)
```

## Results
```{r, eval = run_adaptive_scoring, include = run_adaptive_scoring, results='asis'}
# Adaptive score summary
adaptive_results$directional_score_summary %>%
  reactable(
    sortable = TRUE,
    filterable = TRUE,
    searchable = TRUE
  )

# Top directionally extreme accessions for each variable
top_accessions <- adaptive_results$top_50_extreme_accessions
for (phenotype_name in unique(top_accessions$phenotype)) {
  cat(
    "\n\n## Top directionally extreme accessions:",
    phenotype_name,
    "\n\n"
  )
  
  table_data <- top_accessions %>%
    filter(phenotype == phenotype_name) %>%
    arrange(desc(absolute_directional_score))
  
  print(
    tagList(
      reactable(
        table_data,
        sortable = TRUE,
        filterable = TRUE,
        searchable = TRUE,
        pagination = TRUE,
        defaultPageSize = 10
      )
    )
  )
}
```
# Adaptive Candidate Markers: QTL plot
```{r, eval = run_adaptive_scoring, include = run_adaptive_scoring}
plot_primary_qtl(
  primary_lead_snps = robustness_results$primary_lead_snps,
  bim_file = paste0(qc_prefix, ".bim"),
  direction_table = adaptive_results$snp_directions
)
```

# Biological Interpretation
```{r, eval = run_bio_interp, include = run_bio_interp}
biointerp_df <- run_biointerpretation_workflow(
  config = config,
  robustness_results = robustness_results,
  overwrite_annotation = config$run$overwrite
)
```

## Results
```{r, eval = run_bio_interp, include = run_bio_interp}
biointerp_df %>%
  reactable(
    sortable = TRUE,
    filterable = TRUE,
    searchable = TRUE,
    pagination = TRUE,
    defaultPageSize = 10,
    defaultColDef = colDef(
      format = colFormat(digits = 3)
    )
  )
```

## Candidate Gene Summary
```{r, eval = run_bio_interp, include = run_bio_interp}
candidate_gene_summary <- summarize_candidate_genes(biointerp_df)
  
candidate_gene_summary %>% 
  reactable(
    sortable = TRUE,
    filterable = TRUE,
    searchable = TRUE,
    pagination = TRUE,
    defaultPageSize = 10,
    defaultColDef = colDef(
      format = colFormat(digits = 3)
    )
  )
```
