---
title: "`r yaml::read_yaml('config.yaml')$project$name`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Load functions
function_files <- sort(list.files("Functions", pattern = "\\.R$", full.names = TRUE))
invisible(lapply(function_files, source))

# Load config
config <- yaml::read_yaml("config.yaml")

has_stage <- function(stage) file.exists(pipeline_stage_file(config, stage))
has_files <- function(paths) all(file.exists(paths))

env_result <- tryCatch(load_environment_stage(config), error = function(e) NULL)
qc_result <- tryCatch(load_qc_stage(config), error = function(e) NULL)
kinship_result <- if (has_stage("kinship")) load_kinship_stage(config) else NULL
pca_result <- if (file.exists(config$pca$output_file)) readRDS(config$pca$output_file) else NULL

# GEA file outputs
gemma_files <- file.path(
  config$gemma$output_dir,
  c("gemma_all_results.csv", "gemma_all_evaluation.csv", "gemma_best_strategy_by_var.csv")
  )
lfmm_files <- file.path(
  config$lfmm$output_dir,
  c("lfmm_all_results.csv", "lfmm_all_evaluation.csv", "lfmm_best_strategy_by_var.csv")
  )
rda_files <- file.path(
  config$rda$output_dir,
  c("rda_all_results.csv", "rda_all_evaluation.csv", "rda_best_strategy_by_var.csv")
  )
pcadapt_files <- file.path(
  config$pcadapt$output_dir,
  c("pcadapt_all_results.csv", "pcadapt_all_evaluation.csv", "pcadapt_best_strategy.csv")
  )

# Loaders
gemma_results <- if (has_files(gemma_files)) load_gemma_results(config) else NULL
lfmm_results <- if (has_files(lfmm_files)) load_lfmm_results(config) else NULL
rda_results <- if (has_files(rda_files)) load_rda_results(config) else NULL
pcadapt_results <- if (has_files(pcadapt_files)) load_pcadapt_results(config) else NULL
snp_overlap_results <- if (has_stage("snp_overlap")) load_overlap_stage(config) else NULL
consensus_results <- if (consensus_files_exist(config, config$env$vars)) load_consensus_results(config) else NULL
ld_results <- if (ld_block_files_exist(config, config$env$vars)) load_ld_block_results(config) else NULL

# File outputs
primary_files <- file.path(
  config$ld_pruning_ph$output_dir,
  c("primary_ld_summary.csv", "final_primary_ld_pruned_snps.csv")
  )
primary <- if (has_files(primary_files) && primary_snp_files_match_config(config)) load_primary_snp_results(config) else NULL
adaptive_files <- file.path(
  config$adaptive_scoring$output_dir,
  c(
  "adaptive_snp_direction_table.csv", "adaptive_snp_direction_summary.csv",
  "accession_directional_scores.csv", "top_50_directionally_extreme_accessions_by_variable.csv",
  "top_50_higher_direction_accessions_by_variable.csv", "top_50_lower_direction_accessions_by_variable.csv",
  "directional_score_summary_by_variable.csv"
)
)
adaptive_results <- if (has_files(adaptive_files)) load_adaptive_scoring_results(config) else NULL
bio_file <- file.path(
  config$biological_interpretation$output_dir,
  "mapped_candidate_genes_annotated.rds"
  )
biointerp_df <- if (file.exists(bio_file)) load_biointerpretation_results(config) else NULL
candidate_gene_file <- file.path(
  config$biological_interpretation$output_dir,
  "candidate_gene_summary.rds"
  )
candidate_gene_csv <- file.path(
  config$biological_interpretation$output_dir,
  "candidate_gene_summary.csv"
  )
candidate_gene_summary <- if (file.exists(candidate_gene_file)) {
  load_candidate_gene_summary(config)
} else if (file.exists(candidate_gene_csv)) {
  read.csv(candidate_gene_csv, check.names = FALSE)
} else NULL

# Reactable settings
show_table <- function(x, page_size = 10) {
  reactable::reactable(x, sortable = TRUE, filterable = TRUE, searchable = TRUE,
                       pagination = TRUE, defaultPageSize = page_size)
}
```

# Pipeline Run Summary

This report reads previously saved pipeline results. It does not run any analytical modules.

```{r}
manifest_file <- file.path(
  pipeline_stage_dir(config),
  "pipeline_manifest.csv"
  )

if (file.exists(manifest_file)) {
  show_table(read.csv(manifest_file, check.names = FALSE))
} else {
  cat("No stage manifest was found. Run `Rscript Scripts/run_pipeline.R` before rendering this report.")
}
```

# Environmental Data Processing

```{r, eval=!is.null(env_result), results='asis'}
climate_data <- env_result$climate_data
metadata_env <- env_result$metadata

for (variable in config$env$vars) {
  cat("\n\n## Geographic distribution: ", variable, "\n\n", sep = "")
  print(plot_climate_map(climate_data, metadata_env, config$metadata$sample_col,
                         config$env$lon_col, config$env$lat_col, variable))
}

cat("\n\n## Environmental correlation\n\n")
print(plot_climate_cor_heatmap(climate_data))

cat("\n\n## Climate PCA\n\n")
print(plot_climate_pca(env_result$climate_pca))
```

# Genotype Quality Control

```{r, eval=!is.null(qc_result)}
qc_result$dimensions %>% show_table()
qc_result$marker_summary %>% show_table()
```

```{r, eval=!is.null(qc_result)}
qc_images <- file.path(
  "Output/QC",
  c("QC_genotype_marker_count_by_chr.png",
    "QC_genotype_marker_positions_by_chr.png")
  )
qc_images <- qc_images[file.exists(qc_images)]
if (length(qc_images) > 0) knitr::include_graphics(qc_images)
```

# Kinship

## Summary and diagnostics

```{r, eval=!is.null(kinship_result)}
kinship_result$stats %>% show_table()
kinship_result$diagnostics$summary %>% show_table()

print(kinship_result$diagnostics$plots$diagonal)
print(kinship_result$diagnostics$plots$off_diagonal)
print(kinship_result$diagnostics$plots$max_similarity)
```

## Filtering

```{r, eval=!is.null(kinship_result)}
kinship_result$diagnostics$duplicate_pairs %>% show_table()
subset(kinship_result$diagnostics$diagonal, flagged) %>% show_table()
kinship_result$filter$removal_table %>% show_table()
```

# Population Structure

## PCA

```{r, eval=!is.null(pca_result)}
metadata_pca <- if (!is.null(qc_result)) qc_result$metadata else NULL

plot_pca(pca_result, metadata_pca, config$metadata$sample_col, NULL, 1, 2)
```

## Scree and cumulative variance

```{r, eval=!is.null(pca_result)}
plot_pca_scree(pca_result)
plot_pca_cumulative_variance(pca_result)
```

# Genome Environment Association

## LMM (GEMMA)

```{r, eval=!is.null(gemma_results)}
gemma_results$evaluation %>% show_table()
gemma_results$best_by_variable %>%
  dplyr::select(-dplyr::any_of(c("lambda_distance", "lambda_acceptable", "has_signal"))) %>% show_table()
```

```{r, eval=!is.null(gemma_results), results='asis'}
for (phenotype in gemma_results$best_by_variable$phenotype) {
  cat("\n\n### QQ plot: ", phenotype, "\n\n", sep = "")
  print(plot_gemma_qq(gemma_results$results, phenotype))
}
```

## LFMM

```{r, eval=!is.null(lfmm_results)}
lfmm_results$evaluation %>% dplyr::select(-dplyr::any_of("K")) %>% show_table()
lfmm_results$best_by_variable %>%
  dplyr::select(-dplyr::any_of(c("K", "lambda_distance", "lambda_acceptable", "has_signal"))) %>% show_table()
```

```{r, eval=!is.null(lfmm_results), results='asis'}
for (phenotype in lfmm_results$best_by_variable$phenotype) {
  cat("\n\n### QQ plot: ", phenotype, "\n\n", sep = "")
  print(plot_lfmm_qq(lfmm_results$results, phenotype))
}
```

## RDA

```{r, eval=!is.null(rda_results)}
rda_results$evaluation %>% show_table()
rda_results$best_by_variable %>%
  dplyr::select(-dplyr::any_of(c("lambda_distance", "lambda_acceptable", "has_signal"))) %>% show_table()
```

```{r, eval=!is.null(rda_results), results='asis'}
for (phenotype in rda_results$best_by_variable$phenotype) {
  cat("\n\n### QQ plot: ", phenotype, "\n\n", sep = "")
  print(plot_rda_qq(rda_results$results, phenotype))
}
```

## pcadapt

```{r, eval=!is.null(pcadapt_results)}
pcadapt_results$evaluation %>% dplyr::select(-dplyr::any_of("K")) %>% show_table()
pcadapt_results$best_evaluation %>%
  dplyr::select(-dplyr::any_of(c("K", "lambda_distance", "lambda_acceptable", "has_signal"))) %>% show_table()
plot_pcadapt_qq(pcadapt_results$results)
```

# SNP-level Method Overlap

These intersections are calculated at SNP level before LD processing.

```{r, eval=!is.null(snp_overlap_results), results='asis'}
for (phenotype in names(snp_overlap_results$by_variable)) {
  cat("\n\n## ", phenotype, "\n\n", sep = "")
  print(snp_overlap_results$by_variable[[phenotype]]$overlap_plot)
  print(show_table(snp_overlap_results$by_variable[[phenotype]]$intersection_summary))
}
```

# Consensus SNP Sets

The configured primary set is **`r if (!is.null(primary)) primary$set else config$consensus$primary_set`**.

## Evaluation

```{r, eval=!is.null(consensus_results)}
consensus_results$evaluation %>%
  dplyr::select(-dplyr::any_of(c("median_methods", "median_qval"))) %>% show_table()
```

## SNP tables by environmental variable

```{r, eval=!is.null(consensus_results) && !is.null(primary), results='asis'}
for (phenotype in config$env$vars) {
  result <- consensus_results$by_variable[[phenotype]]
  cat("\n\n### ", phenotype, "\n\n", sep = "")
  cat("#### Environmental consensus",
      if (primary$set == "env_2methods") " — selected primary"
      else "", "\n\n")
  cat("SNPs supported by at least two environmental GEA methods.\n\n")
  print(show_table(result$env_2methods))
  cat("\n\n#### High-confidence consensus",
      if (primary$set == "high_confidence") " — selected primary"
      else "", "\n\n")
  cat("SNPs supported by all three environmental methods,
      or at least two environmental methods + pcadapt.\n\n")
  print(show_table(result$high_confidence))
  cat("\n\n#### Exploratory support\n\n")
  cat("SNPs supported by exactly one environmental method + pcadapt.\n\n")
  print(show_table(result$exploratory_support))
}
```

# Post-Hoc LD Processing

```{r, eval=!is.null(ld_results)}
ld_results$summary %>%
  dplyr::select(-dplyr::any_of(c("n_ld_blocks", "median_min_q"))) %>% show_table()
```

# Primary LD-Pruned SNP Set

## Summary

```{r, eval=!is.null(primary)}
primary$summary %>%
  dplyr::select(-dplyr::any_of(c("n_ld_blocks", "median_min_q"))) %>% show_table()
```

## Lead SNPs across all variables

```{r, eval=!is.null(primary)}
primary$lead_snps %>%
  dplyr::select(-dplyr::any_of(c("methods", "n_env_methods", "min_env_p", "min_env_q",
                                 "consensus_set", "ld_block", "max_r2_in_block"))) %>% show_table()
```

# Adaptive Germplasm Scoring

```{r, eval=!is.null(adaptive_results)}
adaptive_results$directional_score_summary %>% show_table()
```

```{r, eval=!is.null(adaptive_results), results='asis'}
top_accessions <- adaptive_results$top_50_extreme_accessions

for (phenotype in unique(top_accessions$phenotype)) {
  cat("\n\n## Top directionally extreme accessions: ", phenotype, "\n\n", sep = "")
  table_data <- top_accessions %>%
    dplyr::filter(.data$phenotype == .env$phenotype) %>%
    dplyr::select(-dplyr::any_of(c("extremeness_percentile", "scored_snp_fraction"))) %>%
    dplyr::arrange(dplyr::desc(.data$absolute_directional_score))
  print(show_table(table_data))
}
```

## Adaptive candidate-marker QTL plot

```{r, eval=!is.null(primary) && !is.null(adaptive_results) && !is.null(qc_result)}
plot_primary_qtl(primary$lead_snps, paste0(qc_result$qc_prefix, ".bim"),
                 direction_table = adaptive_results$snp_directions)
```

# Biological Interpretation

```{r, eval=!is.null(candidate_gene_summary)}
candidate_gene_summary %>% show_table()
```
